`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-01-28 paper/paper.qmd
edits to tidy paper

---
title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
format:
  #md: default
  html:
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nick Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 3 "
  - name: Daniel Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] (@ref-tt_healthcare) and likely a myriad others we haven't thought of. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to supply a set of geographic point locations and an extent of interest, and calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest.

@@

`traveltime` provides a spatial interface using object classes from the `terra` package `[@terra]`. It accepts points as either `data.frame` or `SpatVector` class objects; extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and return the results as a `SpatRaster` object. The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global available via the `malariaAtlas` package [@pfeffer2018malariaatlas]. Rather than house and manipulate these surfaces independently, `traveltime` relies on `malariaAtlas` to download the surfaces for the extent of interest.

@@

`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global.

@@

Other R packages produce superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations. Furthermore, TravelTime.com requires access keys, paid subscription beyond a short free period, caps queries, which add considerable friction to the use experience.

With `traveltime`, we provide internally complete software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.

# Example workflow: walking home from public transport in Singapore

In this example workflow, we wish to calculate the walking travel time to the nearest mass transit station across the island nation of Singapore --- specifically MRT and LRT stations --- and create a map of this.

@@

-   our area of interest; in this case, a map of Singapore, and
-   our points to calcute travel time from, that is, the locations of Singapore's MRT and LRT stations.

We can download a national-level polygon of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata]. Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`).

```{r get basemap}
#| label: "fig-basemap"
#| fig-cap: "Area of interest: Singapore"
library(terra)
library(geodata)
library(tidyterra)

@@

ggplot() +
  geom_spatvector(
    data = sin
  ) +
  # scale_fill_viridis_c(
  #   option = "A",
  #   direction = -1,
  #   na.value = "transparent"
  # ) +
  scale_fill_manual(na.value = "transparent") +
  theme_minimal()
```

The the `stations` data set included in the `traveltime` package is a 563 \$\*\$ 2 matrix containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata:2024.

@@

So now we have the two items of data we require. The next step is to download a friction surface for our area of interest.

We can pass in our basemap `sin` (@fig-basemap), a `SpatVector`, directly as the `extent`. It would also be possible pass in a `SpatRaster` with extent equal to our area of interest, or specify coordinates either as a vector, matrix, or `SpatExtent`.

We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. Alternatively, we could use `"motor2020"` for motorised travel if that was of interest.

@@

@@

Taking a look at these data, below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points @fig-data. Higher values of friction indicate more time travelling across a given cell.

@@

With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations` and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station.

```{r calculate result}
travel_time_sin <- calculate_travel_time(
  friction_surface = friction_singapore,
  points = stations
)
travel_time_sin
```

We present those results in @fig-result:

@@

Note that the above raster include infinite (`Inf`) values. In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland. These raster cells for these islands appear absent in @fig-result. Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot @fig-result.

@@

The `traveltime` package is immediately suitable to a range of applications where travel to custom locations of interest. But we nonetheless see opportunities to build the package utility into the future through two mechanisms: capability to distribute more friction surfaces, and additional methods to handle large spatial extents.

Firstly, `traveltime` currently has access to friction surfaces for 2020, both at 30 arc-second resolution. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a 5 km cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell.

At the other end of the scale, the package can require large amounts of memory for analyses over large landscapes (e.g. one analyses over Africa required \~ 72 GB RAM). Developing methods to handle large landscapes with less memory would be helpful to make such analyses accessible to those without access to larger computing resources.

>>>>

---
title: "`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations"
format:
  #md: default
  html:
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nick Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 3 "
  - name: Daniel Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@weiss2020global] (@ref-tt_healthcare) and likely a myriad others. Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref]. `{traveltime}` enables a user to supply a set of geographic point locations and an extent of interest, to then calculate the motorised or walking travel time over that extent. The result is a raster of 30 arcsecond resolution (approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator) where the value in each cell is the lowest travel time in minutes to any of the supplied point locations over the extent of interest (e.g., travel time from health care facilities @ref-tt_healthcare from @weiss2020global).

@@

`{traveltime}` provides a spatial interface using object classes from the `terra` package `[@terra]`. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and extents as `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` objects; and returns the result as a `SpatRaster` object. The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global available via the `malariaAtlas` package [@pfeffer2018malariaatlas]. Rather than house and manipulate these surfaces independently, `{traveltime}` relies on `malariaAtlas` to download the surfaces for the extent of interest.

@@

`{traveltime}` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `{traveltime}` package, we suggest citations of the package should be accompanied by citing the underlying work by @weiss2018global; @weiss2020global as well.

@@

Other R packages produce superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the <https://www.TravelTime.com> API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons, which are areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across cells. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repetitive iteration for all combinations of time and points, and additional calculations of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use.

With `{traveltime}`, we provide free and open source software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.

# Example: walking from public transport in Singapore

In this example we wish to calculate the walking travel time from the nearest mass transit station across the island nation of Singapore --- specifically MRT and LRT stations --- and create a map of this.

@@

-   our extent of interest --- in this case a map of Singapore, and
-   our points to calculate travel time from --- here the locations of Singapore's MRT and LRT stations.

We can download a national-level polygon of Singapore from the GADM [@gadm] database using the `geodata` package [@geodata]. Here we download only the national boundary (`level = 0`) and at a low resolution (`resolution = 2`). Our boundary `sin` is a `SpatVector` class object.

```{r get basemap}
library(geodata)

@@

sin
```

```{r plot basemap}
#| label: "fig-basemap"
#| fig-cap: "Area of interest: Singapore"

ggplot() +
  geom_spatvector(
    data = sin
  ) +
  scale_fill_manual(na.value = "transparent") +
  theme_minimal()
```

The the `stations` data set included in the `{traveltime}` package is a 563 \$\*\$ 2 `matrix` containing the longitude (x) and latitude (y) of all LRT and MRT station exits in Singapore from @singdata:2024.

@@

Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin` (@fig-basemap), a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.)

@@

Thus we have our friction surface as a `SpatRaster`:

@@

Below we plot the friction surface raster `friction_singapore`, with the vector boundary of `sin` as a dashed grey line, and `stations` as grey points @fig-data. Higher values of friction indicate more time travelling across a given cell.

@@

With all the data collected, the function `calculate_travel_time` takes the friction surface `friction_singapore` and the locations of interest `stations`, and returns a `SpatRaster` of walking time in minutes to each cell from the nearest station as a `SpatRaster`:

```{r calculate result}
travel_time_sin <- calculate_travel_time(
  friction_surface = friction_singapore,
  points = stations
)

travel_time_sin
```

We present the resulting calculated travel times in @fig-result:

@@

Note that the above raster `travel_time_sin` includes infinite (`Inf`) values. In @fig-data, a number of islands to the south are shown as cells unconnected with the mainland. These raster cells for these islands appear absent in @fig-result. Because they are not connected, the calculated travel time is infinite, and so these cells do not appear in the plot @fig-result.

@@

The `{traveltime}` package is immediately suitable to a range of applications where travel to custom locations of interest. Nonetheless, we see opportunities to build the package utility into the future through two mechanisms: capability to distribute more a wider range friction surfaces, and additional methods to handle large spatial extents.

Firstly, `{traveltime}` currently has access to friction surfaces for 2020, both at 30 arc-second resolution. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a 5 km cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell.

At the other end of the scale, the package can require relatively large amounts of onboard memory for analyses over large landscapes (e.g. one analyses over Africa required \~ 72 GB RAM). Developing methods to handle large landscapes either with less memory or via cloud resources would be helpful to make such analyses accessible to those without access to larger computing resources.
